The Companion Mass Distribution to the ELM WD Sample

### 2014-10-02 documents/letter/ms.tex
experiments

% packages
\usepackage{amssymb,amsmath,amsbsy}

@@

\begin{equation}
	\frac{M_2^3}{\left(M_1+M_2\right)^2} {\rm sin}^3 i = \frac{P_{\rm orb}}{2\pi G} K^3 \label{eq:massfunc}
\end{equation}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.7\columnwidth]{figures/Porb_M1/Porb_M1.jpg}
\caption{The $P_{\rm orb}$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $P_{\rm orb}$ and $M_1$ alone, the two populations are indistinguishable.}
\end{center}
\end{figure}

\section{Methods (?)}

We construct a statistical model to derive constraints on a parametric model, parametrized by the vector $\bs{\theta}$,\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$. For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $m_f$ (Eq.~\ref{eq:massfunc}). We would like to find posterior constraints on the parameters, $\boldsymbol{\theta}$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given m_f) = \frac{1}{\mathcal{Z}}~p(m_f \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(m_f \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(m_f \given \bs{\theta})$, involves integrals over the latent variables $i$ and companion mass, $M_2$,
\begin{equation}
    p(m_f \given \bs{\theta}) = \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2~p(m_f \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{equation}
where we neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles, $i$, are isotropically distributed
\begin{align}
	p(m_f \given M_1, M_2, i) &= \delta(m_f - f(M_1, M_2, i))\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~ \delta(m_f - f(M_1, M_2, i))~\sin i\\
    &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~\sin i ~ \delta(g(M_1,M_2,i))\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - m_f.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta{(G(x)}) = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in Equation~\ref{eq:delta} are 
\begin{align}
	i^* &= \sin^{-1}\alpha\\
	\alpha &= \frac{(m_f(M_1+M_2)^2)^{1/3}}{M_2}\\
	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{\sin i^*}{g'(M_1,M_2,i^*)}\\
	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - (m_f(M_1+M_2)^2)^{2/3}}}\label{eq:fullm2}
\end{align}
where the bottom bound in the integral in Equation~\ref{eq:fullm2} is the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$.

[TODO: put in $p(M_2\given ...), description of models in separate subsection?]

In future work we will extend this analysis to incorporate uncertainties and non-detections.

\section{Our model}

>>>>

% packages
\usepackage{amssymb,amsmath,amsbsy}
\usepackage{booktabs}

@@

\begin{equation}
	m_f = \frac{M_2^3}{\left(M_1+M_2\right)^2} {\rm sin}^3 i = \frac{P_{\rm orb}}{2\pi G} K^3 \label{eq:massfunc}
\end{equation}

@@

%\begin{figure}[h!]
%\begin{center}
%\includegraphics[width=0.7\columnwidth]{figures/Porb_M1/Porb_M1.jpg}
%\caption{The $P_{\rm orb}$ - $M_1$ distribution of the ELM WD sample. The three eclipsing systems, with known $M_2$ are shown by the filled circles. Arrows indicate the masses of ELM WDs without detected radial velocity variation. Triangles indicate the positions of the known WD-NS binaries. From $P_{\rm orb}$ and $M_1$ alone, the two populations are indistinguishable.}
%\end{center}
%\end{figure}

\section{Methods}

We construct a statistical model to derive constraints on a parametric model for the distribution of ELM WD companion masses, $p(M_2 \given \bs{\theta})$.\footnote{In what follows, vectors or sets of parameters or quantites are represented by bold symbols.} For each system, we assume we are given $K$, $T$, and $M_1$ and therefore know the value of the mass function, $m_f$ (Eq.~\ref{eq:massfunc}). We would like to find posterior constraints on the parameters, $\bs{\theta}$, that describe the distribution of companion masses, $p(M_2\given \bs{\theta})$, given all of the observed mass functions, $\bs{m}_f$, by deconvolving the distribution of mass functions from the unobserved inclination angle, $i$. Using Bayes' rule,
\begin{equation}
    p(\bs{\theta} \given m_f) = \frac{1}{\mathcal{Z}}~p(m_f \given \bs{\theta})~p(\bs{\theta}).
\end{equation}
where $p(m_f \given \bs{\theta})$ is the likelihood, $p(\bs{\theta})$ is the prior on parameters $\bs{\theta}$, and the evidence integral, $\mathcal{Z}$, is a constant that depends only on the data. The marginal likelihood, $p(m_f \given \bs{\theta})$, involves integrals over the latent variables $i$ and companion mass, $M_2$,
\begin{equation}
    p(m_f \given \bs{\theta}) = \int_0^{\pi/2} di \int_{-\infty}^\infty dM_2~p(m_f \given M_1, M_2, i)~p(M_2 \given \bs{\theta})~p(i).
\end{equation}
where we neglect observational uncertainties in the mass function,\footnote{This is justified because the relative uncertainties in these quantities are small, $\sigma_x / x \sim 0.1$, however in future work we will extend this methodology to include uncertainties in all measured quantities.} and assume the inclination angles, $i$, are isotropically distributed
\begin{align}
	p(m_f \given M_1, M_2, i) &= \delta(m_f - f(M_1, M_2, i))\\
	f(M_1, M_2, i) &= \frac{(M_2 \sin i)^3}{(M_1 + M_2)^2}\\
	p(i) &= \sin i.
\end{align}
For now, we keep this general by not yet specifying a parametric form for the companion mass distribution, $p(M_2 \given \bs{\theta})$. With the above assumptions, the marginal likelihood integral is
\begin{align}
    p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~ \delta(m_f - f(M_1, M_2, i))~\sin i\\
    &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta}) \int_0^{\pi/2} di ~\sin i ~ \delta(g(M_1,M_2,i))\label{eq:delta}
\end{align}
where
\begin{equation}
	g(M_1,M_2,i) = \frac{M_2^3}{(M_1+M_2)^2}\sin^3 i - m_f.
\end{equation}
The inner integral (over $i$) has the form
\begin{equation}
    \int dx~F(x)~\delta{(G(x)}) = \sum_j \frac{F(x^*_j)}{|G'(x^*_j)|}
\end{equation}
where the sum is over the roots, $x^*_j$, of the function $G(x)$. The root, $i^*$, and derivative of the argument of the delta function in Equation~\ref{eq:delta} are 
\begin{align}
	i^* &= \sin^{-1}\alpha\\
	\alpha &= \frac{(m_f(M_1+M_2)^2)^{1/3}}{M_2}\\
	\frac{\partial g}{\partial i} &= \frac{3M_2^3}{(M_1+M_2)^2}\sin^2 i \cos i\\
	\frac{\partial g}{\partial i}\bigg\rvert_{i^*} &= \frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}
\end{align}
With this in mind, we may write the marginal likelihood as
\begin{align}
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{\sin i^*}{g'(M_1,M_2,i^*)}\\
	&= \int_{0}^\infty dM_2~p(M_2 \given \bs{\theta})~\alpha\left[\frac{3M_2^3}{(M_1+M_2)^2} \alpha^2 \sqrt{1 - \alpha^2}\right]^{-1}\\
	&= \int_{M_{2,{\rm min}}}^\infty dM_2~p(M_2 \given \bs{\theta})~\frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - (m_f(M_1+M_2)^2)^{2/3}}}\label{eq:fullm2}
\end{align}
where the bottom bound in the integral in Equation~\ref{eq:fullm2} is set by the minimum companion mass for which the integrand is real, $M_{2,{\rm min}}$.

\section{Experiments} \label{sec:experiments}

We must now choose a functional form for the companion mass distribution,  $p(M_2\given \bs{\theta})$. For all experiments -- including that with the real data -- we use a two component Gaussian mixture model to fit the data. However, in tests below, we generate test data with a variety of mixture component forms. The likelihood with a Gaussian mixture model is
\begin{align}
	h(M_2, m_f, M_1) &= \frac{(M_1+M_2)^{4/3}}{3m_f^{1/3}M_2\sqrt{M_2^2 - (m_f(M_1+M_2)^2)^{2/3}}}\\
	p(m_f \given \bs{\theta}) &= \int_{0}^\infty dM_2~\left[ (1-f_{\rm NS})p_{\rm WD} + f_{\rm NS}p_{\rm NS} \right]~h(M_2, m_f, M_1)\\
	p_{\rm WD} &= \mathcal{N}(M_2 \given \mu_{\rm WD}, \sigma^2_{\rm WD}) \\
	p_{\rm NS} &= \mathcal{N}(M_2 \given \mu_{\rm NS}, \sigma^2_{\rm NS})
\end{align}
where $\mathcal{N}$ is the normal distribution with mean $\mu$ and variance $\sigma^2$. In this work, we fix the mean and variance of the neutron star distribution to
\begin{align}
	\mu_{\rm NS} &= 1.4~\Msun\\
	\sigma^2_{\rm NS} &= (0.2~\Msun)^2
\end{align}
given the uncertainties in neutron star mass, birth channels, and mass accretion scenarios \citep{refs} [TODO: Jeff, references]. Our companion mass model parameters are then $\bs{\theta} = (\mu_{\rm WD}, \sigma_{\rm WD}, f_{\rm NS})$.

We choose standard priors on these parameters. For the mean of the white dwarf component, we use a uniform distribution from $0.2-1.4~\Msun$. We use a (scale-invariant) logarithmic prior for the variance of the white dwarf component over the range $0.05-2.~\Msun$. We use a uniform distribution over the dimensionless neutron star fraction from $0-1$. The model parameters are summarized in Table~\ref{tbl:parameters}.

\begin{table*}[ht]
\begin{center}
	\begin{tabular}{l c c l} \toprule
		Parameter & Prior \\\toprule
		$f_{\rm NS}$ & $\mathcal{U}(0, 1)$ \\ 
		$\mu_{\rm WD}$ & $\mathcal{U}(0.2, 1.4)$ \\ 
		$\sigma_{\rm WD}$ & $\frac{1}{\ln(40)\sigma}$ $(0.05 < \sigma < 2~\Msun)$ \\ 
		$\mu_{\rm NS}$ & 1.4~$\Msun$ (fixed) \\ 
		$\sigma_{\rm NS}$ &  0.1~$\Msun$ (fixed) \\
		\bottomrule
		\end{tabular}
	\caption{Parameter information for the form of the companion mass distribution used in the experiments of Section~\ref{sec:experiments}. $\mathcal{U}$ is the uniform distribution. There are 3 free parameters. \label{tbl:parameters}}
\end{center}
\end{table*}

\subsection{Single Gaussian (white dwarf)}

\subsection{Two Gaussians (white dwarf + neutron star)}

\subsection{Uniform (white dwarf) + Gaussian (neutron star)}

\hrule{4em}
